Hot and fresh: evidence of climate-related suboptimal water conditions for seagrass in a large Gulf coast estuary

### 2024-06-03 revision2/review_response.qmd

*Response: We have added p-values to the text when describing the relationship of predictors to the seagrass response variables.  Please see section 3.3 for updates.*

For the GAM analysis, you may want to consider removing the interactions with year, to provide more power to define the GAM smoothers.  Most of these interactions are non-significant, and you’ve examined them individually, so you would be justified in removing these. You could mention you did these in preliminary models and no strong patterns were observed.  Or, if there is no difference you could just leave them.  Regardless, it is likely worthwhile to investigate.

*Response: We have removed all interaction terms from the GAMs and updated the figures, tables, and text.  The results were very similar and we agree it was not worth including these in the models.  We have made a note in the text that initial models evaluated interaction terms, but the results were similar.*

*Lines 370-373: "Initial models also included a tensor product interaction term with a cubic regression spline that evaluated the potential interacting effects of each predictor with year. However, most of the interaction terms were not significant and were excluded from the final models."*

@@

Line 491: Change “model for the EPC “ to “seagrass cover…”

*Response: Changed.*

Line 493: seagrass cover was associated with decreased light attenuation for both HB and MTB segments.  In HB, it was a fairly random pattern with not much biological meaning.  For MT though seagrass cover increased with increased light attenuation, but as you mention, this is still below the threshold.  Some clarification here might be useful.

*Response: The following was added on lines 489-493: "Associations with light attenuation were also observed for the HB (p < 0.001) and MTB (p = 0.004) bay segments, although the relationship in HB was flat beyond 0.75 m$^{-1}$ and opposite as expected for MTB (Figure 7, Table S7). As noted above, light attenuation in HB and MTB in recent years is well within the range supportive of seagrass growth (Figure 2c), such that these associations may not be describing meaningful biological relationships."*

@@

Line 498: both metric did not add additional value – however, it is significant for HB.  Not sure if this is relevant or not

*Response: Removing the interaction terms also showed a significant association in OTB, although again this is contrary to expectation as an increase in the metric was associated with an increase in seagrass.  The response to the following comment provides some explanation for this anomaly.*

Line 555: “For the EPC model none of the stressor metrics were associated with seagrass change”.  This isn’t quite correct: – salinity OTB was, and also the general trends in increasing sg with increasing temperature should be acknowledged explicitly. Although these were not significant they do suggest a relationships.  I appreciate the distinction between the temperature metrics used for the different datasets as you outline, but what would the mechanism be?  Could the increase with temperature in the EPC data indicate increased vegetative growth in anomalous temperature conditions?  Several studies have observed this (DuBois et al. 2020, Wong and Dowd 2023 etc)

*Response: Yes, this was an explanation that was initially considered for the temperature metric, but not stated in the original text. It is also difficult to describe the positive association of seagrass change with the salinity metric in OTB.  Our description later in the paragraph regarding the spatial mismatch between the water quality and seagrass data is a possible explanation.*

@@

*The following text was added to the discussion (lines 557-561): "The positive association of seagrass change with the temperature (or 'both') metric could be a signal of increased vegetative growth with temperature anomalies (e.g., Wong and Dowd 2023), whereas the positive association with the salinity metric in OTB is not easily explained since extreme precipitation events have been linked to seagrass loss in Tampa Bay (Greening and Janicki 2006)."*

Line 611: I think this explanation is really important, can it be highlighted above too.  What is the depth difference and does this ameliorate solar heating effects experienced by the seagrass.

*Response: The mean depth of all EPC sampling stations (excluding LTB) is 3.88 meters, whereas the mean depth of points along each transect is 0.76 meters. The following was added to lines 363-365 in section 2.7 of the methods: "...; note that the seagrass transect data were collected at different locations and depths ($\bar{Z}$ = 0.76 m) than the EPC water quality data ($\bar{Z}$ = 3.88 m, excludes LTB not evaluated)."*

### 2024-06-03 manu-draft.qmd
second revision submitted

For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends above. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment from 1998 to 2022. Preliminary analyses evaluated different lagged associations between the stressor metrics and seagrass change, although initial results suggested no additional insight could be gained using lagged assessments compared to the transect year summaries. As such, the stressor metrics were compared to frequency occurrence (all species) each year by bay segment; note that the seagrass transect data were collected at different locations and depths ($\bar{Z}$ = `r sgdep` m) than the EPC water quality data ($\bar{Z}$ = `r epcdep` m, excludes LTB not evaluated). GAMs were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, or both. Additional predictors included year and light attenuation as estimated from Secchi depth [@tbep0696]. For 1998 to 2022, `r onb$epc` of the Secchi observations were recorded on the bottom. A single smooth term was used for each predictor using a thin plate regression spline. Initial models also included a tensor product interaction term with a cubic regression spline that evaluated the potential interacting effects of each predictor with year. However, most of the interaction terms were not significant and were excluded from the final models. Two models were evaluated, one with the bottom temperature and salinity metrics together and another with the both metric. All models excluded Lower Tampa Bay because of minimal seagrass change over time. Additionally, total seagrass frequency occurrence was used as a response variable as an aggregate measure of community change to potential stressors given the likelihood that species-specific changes may be more difficult to model and that the majority of seagrass loss in recent years was dominated by *H. wrightii* in the upper bay segments (Figure [-@fig-seagrasschg]b). All GAMs were fit using restricted maximum likelihood evaluation [@wood11] with the *mgcv* R package \[@wood17.

@@

GAMs to assess the potential effects of year, light attenuation, temperature, and salinity on seagrass change provided some evidence that seagrass change was influenced by climate stressors, particularly for the FIM and PDEM datasets. The seagrass cover data (`r sgmodtxt$epcmod1`, EPC model) had notable associations with year for all bay segments (OTB: `r epc1pvl[['yr:bay_segmentOTB']]`, HB: `r epc1pvl[['yr:bay_segmentHB']]`, MTB: `r epc1pvl[['yr:bay_segmentMTB']]`) that followed long-term changes in @fig-seagrasschg. Associations with light attenuation were also observed for the HB (`r epc1pvl[['la:bay_segmentHB']]`) and MTB (`r epc1pvl[['la:bay_segmentMTB']]`) bay segments, although the relationship in HB was flat beyond 0.75 m$^{-1}$ and opposite as expected for MTB (@fig-sgmod1, Table S7). As noted above, light attenuation in HB and MTB in recent years is well within the range supportive of seagrass growth (Figure [-@fig-seagrasschg]c), such that these associations may not be describing meaningful biological relationships. Associations of seagrass change with the temperature (OTB: `r epc1pvl[['temp:bay_segmentOTB']]`, HB `r epc1pvl[['temp:bay_segmentHB']]`, MTB `r epc1pvl[['temp:bay_segmentMTB']]`) and salinity (OTB: `r epc1pvl[['sal:bay_segmentOTB']]`, HB `r epc1pvl[['sal:bay_segmentHB']]`, MTB `r epc1pvl[['sal:bay_segmentMTB']]`) metrics were not present or contrary to expectation (e.g., increase in seagrass with an increasing salinity metric in OTB, slight increase in seagrass cover with an increasing temperature metric in all bay segments). Similarly, the EPC model using the "both" stressor metric (`r sgmodtxt$epcmod2`, Figure S7, Table S8) had positive associations of the "both" metric with seagrass change in OTB (`r epc2pvl[['both:bay_segmentOTB']]`) and HB (`r epc2pvl[['both:bay_segmentHB']]`).

>>>>

For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends above. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment from 1998 to 2022. Preliminary analyses evaluated different lagged associations between the stressor metrics and seagrass change, although initial results suggested no additional insight could be gained using lagged assessments compared to the transect year summaries. As such, the stressor metrics were compared to frequency occurrence (all species) each year by bay segment; note that the seagrass transect data were collected at different locations and depths ($\bar{Z}$ = `r sgdep` m) than the EPC water quality data ($\bar{Z}$ = `r epcdep` m, excludes LTB not evaluated). GAMs were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, or both. Additional predictors included year and light attenuation as estimated from Secchi depth [@tbep0696]. For 1998 to 2022, `r onb$epc` of the Secchi observations were recorded on the bottom. A single smooth term was used for each predictor using a thin plate regression spline. Initial models also included a tensor product interaction term with a cubic regression spline that evaluated the potential interacting effects of each predictor with year. However, most of the interaction terms were not significant and were excluded from the final models. Two models were evaluated, one with the bottom temperature and salinity metrics together and another with the both metric. All models excluded Lower Tampa Bay because of minimal seagrass change over time. Additionally, total seagrass frequency occurrence was used as a response variable as an aggregate measure of community change to potential stressors given the likelihood that species-specific changes may be more difficult to model and that the majority of seagrass loss in recent years was dominated by *H. wrightii* in the upper bay segments (Figure [-@fig-seagrasschg]b). All GAMs were fit using restricted maximum likelihood evaluation [@wood11] with the *mgcv* R package \[@wood17\].

@@

GAMs to assess the potential effects of year, light attenuation, temperature, and salinity on seagrass change provided some evidence that seagrass change was influenced by climate stressors, particularly for the FIM and PDEM datasets. The seagrass cover data (`r sgmodtxt$epcmod1`, EPC model) had notable associations with year for all bay segments (OTB: `r epc1pvl[['yr:bay_segmentOTB']]`, HB: `r epc1pvl[['yr:bay_segmentHB']]`, MTB: `r epc1pvl[['yr:bay_segmentMTB']]`) that followed long-term changes in @fig-seagrasschg. Associations with light attenuation were also observed for the HB (`r epc1pvl[['la:bay_segmentHB']]`) and MTB (`r epc1pvl[['la:bay_segmentMTB']]`) bay segments, although the relationship in HB was flat beyond 0.75 m$^{-1}$ and opposite as expected for MTB (@fig-sgmod1, Table S7). As noted above, light attenuation in HB and MTB in recent years is well within the range supportive of seagrass growth (Figure [-@fig-seagrasschg]c), such that these associations may not be describing meaningful biological relationships. Associations of seagrass change with the temperature (OTB: `r epc1pvl[['temp:bay_segmentOTB']]`, HB: `r epc1pvl[['temp:bay_segmentHB']]`, MTB: `r epc1pvl[['temp:bay_segmentMTB']]`) and salinity (OTB: `r epc1pvl[['sal:bay_segmentOTB']]`, HB: `r epc1pvl[['sal:bay_segmentHB']]`, MTB: `r epc1pvl[['sal:bay_segmentMTB']]`) metrics were not present or contrary to expectation (e.g., increase in seagrass with an increasing salinity metric in OTB, slight increase in seagrass cover with an increasing temperature metric in all bay segments). Similarly, the EPC model using the "both" stressor metric (`r sgmodtxt$epcmod2`, Figure S7, Table S8) had positive associations of the "both" metric with seagrass change in OTB (`r epc2pvl[['both:bay_segmentOTB']]`) and HB (`r epc2pvl[['both:bay_segmentHB']]`).
